---
title: "Playing Around with Winnipeg Transit Data"
author: "Marcello Nesca"
date: "July 2 2019"
output: github_document
---

## Motivation(s) & Objectives

The idea of this practice dataset is:  
  1) to upkeep my data science skills in R  
  2) playing with version control using Github  
  3) Store my thoughts on data analysis  
  4) i want to practice more on Tidyverse code and stop using BASE R coding  


```{r loading packages, include=FALSE, cache=FALSE}
library(here) 
library(tidyverse)
library(stringr)
library(lubridate)
knitr::opts_chunk$set(fig.width = 12, fig.height = 8)
```
Here is my initial startup for all the packages and options im running.  
library(here)   
library(tidyverse)  
library(stringr)  
library(lubridate)  
knitr::opts_chunk$set(fig.width = 12, fig.height = 8)  

## Loading Datasets

I have obtained this dataset from the City of Winnipeg open dataset archive I will be concentrating on utilizing the Transit-passups for this practice project.  Since this dataset gets updated daily, I have gotten this dataset in June 29th 2019.

UPDATE: 08-24-2019: I had accidently deleted my data file somehow, and so I re-downloaded a new dataset on August 22nd 2019. 

Ref: <https://data.winnipeg.ca/Transit/Transit-Pass-ups/mer2-irmb>

```{r loading data, include=FALSE}
transitdata <- read_csv(here::here("data", "Transit_Pass-ups.csv"))
```

## Data Cleaning and Checking Variables (AKA - preprocessing)

Here I want to describe what my variables initially look like before further analysis. We first need to see what variables are included without continiously looking at the dataframe.  The idea is to try to be efficient in your work.  

UPDATE 07-31-2019:
definitely by practice, cleaning and visualizing data is a circular process. 

Here are a list of the variables uncleaned: 
```{r sumvar}
sumvar <- summary(transitdata)
print(sumvar)
```

### Renaming, and changing variables to factors
```{r Mutating and renaming, include=TRUE}
transitdata <- transitdata %>% 
  dplyr::rename(RouteDestination = `Route Destination`) %>%
  dplyr::rename(RouteName = `Route Name`) %>%
  dplyr::rename(PassUpType = `Pass-Up Type`) %>%
  dplyr::rename(RouteNumber = `Route Number`) %>%
  dplyr::rename(PassUpID = `Pass-Up ID`) %>%
  mutate_at(vars("RouteDestination", "RouteName", "PassUpType", "RouteNumber"), as.factor)
```


#### Testing the renamed variables
```{r GGPLOT Test, include=TRUE}
ggplot(transitdata) +
  geom_bar(aes(PassUpType, fill = PassUpType)) +
  xlab("Type of bus that passed")
```

### Changed longitude and latitude
For potential spatial analysis in the future
```{r Separating variables!, include=TRUE}
transitdata <- transitdata %>%
  separate(Location, into = c("Extra", "Lat1", "Long1"), sep = " ") %>%
  separate(Lat1, into = c("Extra1", "Latitude"), sep = 1) %>%
  separate(Long1, into = c("Longitude", "Extra2"), sep = -1) %>%
  select(-Extra, -Extra1, -Extra2)
```

### splitting up time and date
Here is where we clean the time variable - i want to lubridate the time variable first, then split up time and date
```{r Mutating Time, include=TRUE}
transitdata <- transitdata %>%
  mutate(Time = mdy_hms(Time)) %>%
  separate(Time, into = c("Date", "Clock"), sep = " ")
```

Creating new variables for splitting up dates into months days and years
```{r Changing Dates, include=TRUE}
transitdata <- transitdata %>%
  mutate(year = year(Date), 
         month = month(Date), 
         day = day(Date),
         ) %>%
  mutate_at(vars("year", "month", "day"), as.factor) 
```

UPDATE 08-10-2019:  
some possible solutions for categorical splitting: 
1) derive only the top 10 and bottom 10 passups so filter by most frequent and least frequent  
2) create a category for day of the week "Monday tuesday etc"  
3) create a category for subgroup by region in winnipeg  
4) create a category for what type of speed bus "super express, rapid transit, etc"

```{r adding categories for bus speed, include=TRUE}
transitdata <- transitdata %>%
  mutate(bus_speed = ifelse(RouteNumber %in% c(137,160,161,162,163,170,180,181,183,185), 'RapidTransit',
                            ifelse(RouteNumber %in% c(101,102,109,110), 'Dart',
                                   ifelse(RouteNumber %in% c(1,2,3), 'DowntownSpirit',
                                          ifelse(RouteNumber %in% c(21,22,24,28,30,31,32,40,41,42,46,48,54,57,58,59,64,65,67), 'Express',
                                                 ifelse(RouteNumber %in% c(25,34,35,36), 'SuperExpress', 'Regular'))))))
  
```

number 4 check! Now lets visualize!

```{r creating day of week variable}
transitdata <- transitdata %>%
  mutate(dayofweek = wday(Date, label = TRUE, abbr = FALSE))
```

number 2 check!

It seems that I have cleaned everything I wish to clean, I have separated all the variables in their component parts including the dates! So now lets explore some data finally!

## Exploratory Analysis (Mostly Univariate Analysis on most variables)

A ton of these visualizations are primarily hypothesis generating, but also for the most part for my own practice sake.

```{r include=TRUE}
ggplot(transitdata) +
  geom_bar(aes(year, fill = year)) +
  xlab("Year")
```

Wow! Interesting exploratory information! The rate of pass-ups are increasing per year since 2015! however in 2015 it did go down from 2014. 2019 is a half year since i downloaded the data end of June 2019.

questions derived from this analysis (either answerable or unanswerable via data):  
1) what caused the increase in pass ups from 2014-2018?  
2) is it possible to join another table for extra insights, if so, what would be the common id?

```{r include=TRUE}
ggplot(transitdata) +
  geom_bar(aes(month, fill = month)) +
  xlab("Month")
```

and this is clearly very interesting! most of the passups occur when -- you guessed it! school is starting!

```{r include=TRUE}
transitdata %>%
  ggplot(aes(fct_infreq(day), fill = day)) +
    geom_bar() +
    xlab("Day")

transitdata %>%
  ggplot(aes(dayofweek, fill = dayofweek)) +
  geom_bar() +
  facet_wrap(~ month, nrow = 4) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  xlab("Day of the Week")
```

```{r}
transitdata %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(freq)) +
    geom_histogram(binwidth = 9)

transitdata %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 2500, freq < 3750, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(freq)) +
    geom_histogram(binwidth = 1)

transitdata %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 5000, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(fct_explicit_na(RouteNumber)))) +
    geom_bar()

transitdata %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq < 100, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(fct_explicit_na(RouteNumber)))) +
    geom_bar()

transitdata %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 2500, freq < 3750, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(RouteNumber))) +
    geom_bar()
```

```{r checking new variable i created, include=TRUE}
transitdata %>%
ggplot(aes(fct_infreq(bus_speed), fill = bus_speed)) +
  geom_bar() +
  facet_wrap(~ year, nrow = 4) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  xlab("Type of bus speed") +
  ylab("count of passups")

# two categorical variables - dotmap by year
transitdata %>%
ggplot(aes(year, bus_speed)) +
  geom_count()

# two categorical variables - heatmap by month
transitdata %>% 
  count(month, bus_speed) %>%  
  ggplot(aes(month, bus_speed)) +
    geom_tile(aes(fill = n))

#Finding out which busses are the culprits! 
transitdata %>%
  filter(bus_speed == 'Regular') %>%
  ggplot(aes(fct_infreq(RouteNumber))) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdata %>%
  filter(bus_speed == 'Express') %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +  
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdata %>%
  filter(bus_speed %in% c('SuperExpress')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdata %>%
  filter(bus_speed %in% c('RapidTransit')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdata %>%
  filter(year == '2018') %>%
  ggplot(aes(month)) +
    geom_bar() +
    xlab("Month for current year")

transitdata %>%
  filter(bus_speed %in% c('RapidTransit')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ month, nrow = 4) +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("RouteNumber")

transitdata %>%
  filter(year %in% c('2017', '2018', '2019'), 
         RouteNumber %in% c('11'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(RouteDestination, fill = RouteDestination)) +
    geom_bar() +
    facet_wrap(~ year) +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("Route Destination -- RouteNumber = 11")
# Possible Research Question:  Does a Route Number with multiple destinations(162,11,75) cause more passups than dedicated route busses(160)?


transitdata %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2017', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")

transitdata %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2018', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")

transitdata %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2019', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")
```

## Lessons Learned and Notes

This section will display all the errors, and difficulties I have struggled with. 

### No Spaces in variable names

If there is a space in the variable name, it will cause an error. I initially thought its because i did not change the variables into factors (which was also true) but it was not the cause of the error, since it did not fix it.

```{r eval=FALSE}
ggplot(transitdata) +
  geom_bar(aes("Pass-Up Type")) +
  xlab("Type of bus that passed")
```
>"It seems this produced an error that I am trying to fix, the error in this case is that its outputting the total count of the variable under Route Destination which is not useful at all. I will try to fix this.  
>
>The issue was that the variables were characters to which i have to change to factors.
>
>AAAAND... urgggg! It seems like transforming the variables didnt work either!!"

>UPDATE: 07-21-2019:
So I finally figured out why my GGPLOT geom bar is continiously breaking -- it took me a week or so to figure this out.  It is because GGPLOT *CANNOT* have spaces in the variables.  Furthermore, spaces between variables is not good coding practice anyways, so... City of Winnipeg, do *NOT* put spaces in your variables.

### Sorting categorical variables by descending order

I struggled with this for the long time, trying to find a tidyverse way of coding of sorting categories by decending order.

>I cannot seem to find a way to properly sort categorical variables but ill definitely find out by next time! There are way too many values for these variables... Now a new problem to solve!

>08-22-2019: I learned something new! forcats package is for all categorical variables... in this case i learned how to sort categorical variables by descending frequency.

### Variables with a ton of categories

I am still currently having an issue on how to deal with variables with a ton of categories. I am trying to find a way to get top 10 results. 

```{r include=TRUE}
ggplot(transitdata) +
  geom_bar(aes(RouteDestination)) +
  coord_flip() +
  xlab("Where the bus went")

ggplot(transitdata) +
  geom_bar(aes(RouteName)) +
  coord_flip() +
  xlab("Name of the Route")
```
