---
title: "Playing Around with Transit Data"
author: "Marcello Nesca"
date: "July 2 2019"
output: github_document
---

## Motivation(s) & Objectives

The idea of this practice dataset is:  
  1) to upkeep my data science skills in R  
  2) playing with version control using Github  
  3) Store my thoughts on data analysis  
  4) i want to practice more on Tidyverse code and stop using BASE R coding  


```{r loading packages, include=FALSE}
library(here) 
library(tidyverse)
library(stringr)
library(lubridate)
```

## Loading Datasets

I have obtained this dataset from the City of Winnipeg open dataset archive I will be concentrating on utilizing the Transit-passups for this practice project.  Since this dataset gets updated daily, I have gotten this dataset in June 29th 2019.

Ref: <https://data.winnipeg.ca/Transit/Transit-Pass-ups/mer2-irmb>

```{r loading data, include=TRUE}
transitdata <- read_csv(here::here("data", "Transit_Pass-ups.csv"))
```

## Data Cleaning and Checking Variables

Here I want to describe what my variables initially look like before further analysis. We first need to see what variables are included without continiously looking at the dataframe.  The idea is to try to be efficient in your work. 
Here are a list of the variables: 
```{r sumvar}
sumvar <- summary(transitdata)
print(sumvar)
```

```{r Descriptive Analysis, include=TRUE}
g <- ggplot(transitdata)
g + 
  geom_bar(aes("Pass-Up Type")) +
  xlab("Type of bus that passed")
```

It seems this produced an error that I am trying to fix, the error in this case is that its outputting the total count of the variable under Route Destination which is not useful at all. I will try to fix this.

The issue was that the variables were characters to which i have to change to factors.

AAAAND... urgggg! It seems like transforming the variables didnt work either!!


```{r Mutating character variables to Factors and renaming, include=TRUE}
transitdata <- transitdata %>% 
  mutate_at(vars("Route Destination", "Route Name", "Pass-Up Type"), as.factor)

## I know this next bit of code is inefficient so i have to trouble shoot it later.
transitdata <- transitdata %>% 
  dplyr::rename(RouteDestination = `Route Destination`) %>%
  dplyr::rename(RouteName = `Route Name`) %>%
  dplyr::rename(PassUpType = `Pass-Up Type`) %>%
  dplyr::rename(RouteNumber = `Route Number`) %>%
  dplyr::rename(PassUpID = `Pass-Up ID`)
tibble(transitdata)
```

UPDATE: 07-21-2019:
So I finally figured out why my GGPLOT geom bar is continiously breaking -- it took me a week or so to figure this out.  It is because GGPLOT *CANNOT* have spaces in the variables.  Furthermore, spaces between variables is not good coding practice anyways, so... City of Winnipeg, do *NOT* put spaces in your variables.

```{r GGPLOT Test, include=TRUE}
g <- ggplot(transitdata)
g + 
  geom_bar(aes(PassUpType, fill = PassUpType)) +
  xlab("Type of bus that passed")
```

```{r Now lets see other variables!, include=TRUE}
g + 
  geom_bar(aes(RouteDestination)) +
  coord_flip() +
  xlab("Where the bus went")

g + 
  geom_bar(aes(RouteName)) +
  coord_flip() +
  xlab("Name of the Route")
```

There are way too many values for these variables... Now a new problem to solve!

UPDATE 07-31-2019:
definitely by practice, cleaning and visualizing data is a circular process. 

```{r Separating variables!, include=TRUE}
transitdata <- transitdata %>%
  separate(Location, into = c("Extra", "Lat1", "Long1"), sep = " ") %>%
  separate(Lat1, into = c("Extra1", "Latitude"), sep = 1) %>%
  separate(Long1, into = c("Longitude", "Extra2"), sep = -1) %>%
  select(-Extra, -Extra1, -Extra2)
tibble(transitdata)
```



```{r Changing Dates, include=TRUE}
## Error doesnt work
transitdata <- transitdata %>%
  mutate(year = year(Time), 
         month = month(Time), 
         day = day(Time),
         hour = hour(Time),
         minute = minute(Time),
         second = second(Time)
         )
```

