---
title: "Playing Around with Winnipeg Transit Data"
author: "Marcello Nesca"
date: "July 2 2019"
output:
  word_document: default
  html_document: default
---

## Motivation(s) & Objectives

The idea of this practice dataset is:  
  1) to upkeep my data science skills in R  
  2) playing with version control using Github  
  3) Store my thoughts on data analysis  
  4) i want to practice more on Tidyverse code and stop using BASE R coding  

Here is my initial startup for all the packages and options im running.  

```{r loading packages}
library(here) 
library(tidyverse)
library(stringr)
library(hms)
library(knitr)
library(reprex)
library(datapasta)
opts_chunk$set(fig.width = 12, fig.height = 8)
```


## Loading Datasets

I have obtained this dataset from the City of Winnipeg open dataset archive I will be concentrating on utilizing the Transit-passups for this practice project.  Since this dataset gets updated daily, I have gotten this dataset in June 29th 2019.

UPDATE: 08-24-2019: I had accidently deleted my data file somehow, and so I re-downloaded a new dataset on August 22nd 2019. 

Ref: <https://data.winnipeg.ca/Transit/Transit-Pass-ups/mer2-irmb>

```{r loading data}
transitdata <- read_csv(here::here("data", "Transit_Pass-ups.csv"))
summary(transitdata)
```

## Data Cleaning and Checking Variables (AKA - preprocessing)

Here I want to describe what my variables initially look like before further analysis. We first need to see what variables are included without continiously looking at the dataframe.  The idea is to try to be efficient in your work.  

UPDATE 07-31-2019:
definitely by practice, cleaning and visualizing data is a circular process. 

### Renaming, and changing variables to factors
```{r Mutating and renaming}
transitdataclean <- transitdata %>% 
  rename(RouteDestination = `Route Destination`) %>%
  rename(RouteName = `Route Name`) %>%
  rename(PassUpType = `Pass-Up Type`) %>%
  rename(RouteNumber = `Route Number`) %>%
  rename(PassUpID = `Pass-Up ID`) %>%
  mutate_at(vars("RouteDestination", "RouteName", "PassUpType", "RouteNumber"), as.factor) # %>%
  # count(fct_explicit_na(RouteNumber), fct_explicit_na(RouteDestination), sort = TRUE)
```

### splitting up time and date
Here is where we clean the time variable - i want to lubridate the time variable first, then split up time and date

```{r hardcoding starttime}
starttime <- hms::hms(00, 30, 05)
```


```{r Mutating Time}
transitdataclean <- transitdataclean %>%
  mutate(Time = lubridate::mdy_hms(Time)) %>%
  separate(Time, into = c("Date", "Clock"), sep = " ") %>%
  mutate(Clock_s = as_hms(Clock)) %>%
  mutate_at(vars("Clock_s"), as.numeric)
```

This next line of code is to subtract the time in which the bus passed up to the 'assumed' start time of 05:30 which is when busses get issued. As a result it counts the amount of seconds from 5:30am to when the bus passed up. The reason why i did this is because i want to see on average what time busses passup and which routes passup the most at certain times during the day.  

I of course expect rush hour but the data should see where this truly lies.
```{r creating time to pass up}
transitdataclean <- transitdataclean %>%
  mutate(timetopassup = as_hms(Clock) - as_hms(starttime)) %>%
  mutate_at(vars("timetopassup"), as.numeric)
```

```{r}
library(tidyverse)
transitdataclean <- transitdataclean %>%
  mutate(passuphms = hms::hms(timetopassup))
```


Creating new variables for splitting up dates into months days and years
```{r Changing Dates}
transitdataclean <- transitdataclean %>%
  mutate(year = lubridate::year(Date), 
         month = lubridate::month(Date), 
         day = lubridate::day(Date),
         ) %>%
  mutate_at(vars("year", "month", "day"), as.factor) 
```

UPDATE 08-10-2019:  
some possible solutions for categorical splitting: 
1) derive only the top 10 and bottom 10 passups so filter by most frequent and least frequent  
2) create a category for day of the week "Monday tuesday etc"  
3) create a category for subgroup by region in winnipeg  
4) create a category for what type of speed bus "super express, rapid transit, etc"

```{r adding categories for bus speed}
transitdataclean <- transitdataclean %>%
  mutate(bus_speed = ifelse(RouteNumber %in% c(137,160,161,162,163,170,180,181,183,185), 'RapidTransit',
                            ifelse(RouteNumber %in% c(101,102,109,110), 'Dart',
                                   ifelse(RouteNumber %in% c(1,2,3), 'DowntownSpirit',
                                          ifelse(RouteNumber %in% c(21,22,24,28,30,31,32,40,41,42,46,48,54,57,58,59,64,65,67), 'Express',
                                                 ifelse(RouteNumber %in% c(25,34,35,36), 'SuperExpress', 'Regular'))))))
## number 4 check!
```

```{r creating day of week variable}
transitdataclean <- transitdataclean %>%
  mutate(dayofweek = lubridate::wday(Date, label = TRUE, abbr = FALSE))
## number 2 check!
```



It seems that I have cleaned everything I wish to clean, I have separated all the variables in their component parts including the dates! So now lets explore some data finally!

## Exploratory Analysis (Mostly Univariate Analysis on most variables)

A ton of these visualizations are primarily hypothesis generating, but also for the most part for my own practice sake.

```{r year}
library(tidyverse)
transitdataclean %>%
ggplot(aes(year, fill = year)) +
  geom_bar() +
  xlab("Year")
```

Wow! Interesting exploratory information! The rate of pass-ups are increasing per year since 2015! however in 2015 it did go down from 2014. 2019 is a half year since i downloaded the data end of June 2019.

questions derived from this analysis (either answerable or unanswerable via data):  
1) what caused the increase in pass ups from 2014-2018?  
2) is it possible to join another table for extra insights, if so, what would be the common id?

```{r}
transitdataclean %>%
ggplot(aes(month, fill = month)) +
  geom_bar() +
  xlab("Month")
```

and this is clearly very interesting! most of the passups occur when -- you guessed it! school is starting!

```{r day}
transitdataclean %>%
  ggplot(aes(fct_infreq(day), fill = day)) +
    geom_bar() +
    xlab("Day")

transitdataclean %>%
  ggplot(aes(dayofweek, fill = dayofweek)) +
  geom_bar() +
  facet_wrap(~ month, nrow = 4) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  xlab("Day of the Week")
```

```{r ggplot practice}
transitdataclean %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(freq)) +
    geom_histogram(binwidth = 9)

transitdataclean %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 2500, freq < 3750, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(freq)) +
    geom_histogram(binwidth = 1)

transitdataclean %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 5000, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(fct_explicit_na(RouteNumber)))) +
    geom_bar()

transitdataclean %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq < 100, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(fct_explicit_na(RouteNumber)))) +
    geom_bar()

transitdataclean %>% 
  group_by(fct_explicit_na(RouteNumber)) %>%
  mutate(freq = n()) %>%
  filter(freq > 2500, freq < 3750, 
         PassUpType == 'Full Bus Pass-Up',
         year == '2018') %>%
  print(transitdata) %>%
  ggplot(aes(fct_infreq(RouteNumber))) +
    geom_bar()
```

```{r more ggplot practice}
transitdataclean %>%
ggplot(aes(fct_infreq(bus_speed), fill = bus_speed)) +
  geom_bar() +
  facet_wrap(~ year, nrow = 4) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  xlab("Type of bus speed") +
  ylab("count of passups")

# two categorical variables - dotmap by year
transitdataclean %>%
ggplot(aes(year, bus_speed)) +
  geom_count()

# two categorical variables - heatmap by month
transitdataclean %>% 
  count(month, bus_speed) %>%  
  ggplot(aes(month, bus_speed)) +
    geom_tile(aes(fill = n))

#Finding out which busses are the culprits! 
transitdataclean %>%
  filter(bus_speed == 'Regular') %>%
  ggplot(aes(fct_infreq(RouteNumber))) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdataclean %>%
  filter(bus_speed == 'Express') %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +  
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdataclean %>%
  filter(bus_speed %in% c('SuperExpress')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdataclean %>%
  filter(bus_speed %in% c('RapidTransit')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar(na.rm = TRUE) +
    xlab("RouteNumber")

transitdataclean %>%
  filter(year == '2019') %>%
  ggplot(aes(month)) +
    geom_bar() +
    xlab("Month for current year")

transitdataclean %>%
  filter(bus_speed %in% c('RapidTransit')) %>%
  ggplot(aes(RouteNumber, fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ month, nrow = 4) +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("RouteNumber")

transitdataclean %>%
  filter(year %in% c('2017', '2018', '2019'), 
         RouteNumber %in% c('11'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(RouteDestination, fill = RouteDestination)) +
    geom_bar() +
    facet_wrap(~ year) +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("Route Destination -- RouteNumber = 11")
# Possible Research Question:  Does a Route Number with multiple destinations(162,11,75) cause more passups than dedicated route busses(160)?

# Another Research Question: Is there a difference between time to event for busses?

transitdataclean %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2017', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")

transitdataclean %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2018', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")

transitdataclean %>% 
  group_by(RouteNumber) %>% 
  mutate(freq = n()) %>% 
  ungroup() %>% 
  filter(freq > 2000, year == '2019', PassUpType == 'Full Bus Pass-Up') %>%
  select(-freq) %>%
  ggplot(aes(fct_infreq(RouteNumber), fill = RouteNumber)) +
    geom_bar() +
    facet_wrap(~ year) +
    xlab("RouteNumber")
```

```{r checking september data yoy, include=TRUE, eval=TRUE}
transitdataclean %>%
  filter(year %in% c('2017', '2018', '2019'), 
         month %in% c('9'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(month, fill = month)) +
    geom_bar() +
    facet_wrap(~ year) +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("September -- Year by Year")
```

```{r}
transitdataclean %>%
  filter(year %in% c('2017', '2018', '2019'), 
         RouteNumber %in% c('11', '162', '160', '75'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(RouteNumber, timetopassup, fill = RouteNumber)) +
    geom_boxplot() +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("Route Numbers")
```

```{r}
transitdataclean %>%
  filter(year %in% c('2018'), 
         bus_speed %in% c('RapidTransit'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(RouteNumber, timetopassup, fill = RouteNumber)) +
    geom_boxplot() +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("Route Numbers")
```

```{r}
transitdataclean %>%
  filter(year %in% c('2018'), 
         RouteDestination %in% c('To University of Manitoba'),
         PassUpType == 'Full Bus Pass-Up') %>%
  ggplot(aes(RouteNumber, hms(Clock_s), fill = RouteNumber)) +
    geom_boxplot() +
    theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    xlab("Route Numbers")
```

## Lessons Learned and Notes

This section will display all the errors, and difficulties I have struggled with. 

### No Spaces in variable names

If there is a space in the variable name, it will cause an error. I initially thought its because i did not change the variables into factors (which was also true) but it was not the cause of the error, since it did not fix it.

```{r}
ggplot(transitdata) +
  geom_bar(aes("Pass-Up Type")) +
  xlab("Type of bus that passed")
```
"It seems this produced an error that I am trying to fix, the error in this case is that its outputting the total count of the variable under Route Destination which is not useful at all. I will try to fix this.  

The issue was that the variables were characters to which i have to change to factors.

AAAAND... urgggg! It seems like transforming the variables didnt work either!!"

UPDATE: 07-21-2019:
So I finally figured out why my GGPLOT geom bar is continiously breaking -- it took me a week or so to figure this out.  It is because GGPLOT *CANNOT* have spaces in the variables.  Furthermore, spaces between variables is not good coding practice anyways, so... City of Winnipeg, do *NOT* put spaces in your variables.

### Sorting categorical variables by descending order

I struggled with this for the long time, trying to find a tidyverse way of coding of sorting categories by decending order.

I cannot seem to find a way to properly sort categorical variables but ill definitely find out by next time! There are way too many values for these variables... Now a new problem to solve!

08-22-2019: I learned something new! forcats package is for all categorical variables... in this case i learned how to sort categorical variables by descending frequency.

### Variables with a ton of categories

I am still currently having an issue on how to deal with variables with a ton of categories. I am trying to find a way to get top 10 results. 

```{r}
#ggplot(transitdata) +
#  geom_bar(aes(RouteDestination)) +
#  coord_flip() +
# xlab("Where the bus went")

#ggplot(transitdata) +
#  geom_bar(aes(RouteName)) +
#  coord_flip() +
#  xlab("Name of the Route")
```

### Learning how to use Reprex() and preparing my data for reprex using datapasta

Today (10-22-2019) i learned something very valuable and one of the reasons i am currently really happy about this is because, when in doubt i can join the RStudio Community and ask for help in a very user friendly way for the helpers and not only that whats important is that i can do this on my own. 

I cannot stress enough how empowering this is for a person learning new code. Most of the help i usually go to are books, workshops, or even my school colleagues or teachers, and sometimes that is not enough for a particular context for *my* code. so to have an entire world community helping me, i can feel more confident with my coding! 

the packages to download are as follows:

```{r Packages for reprex and datapasta}
#install.packages(c("reprex", "datapasta"))

#library(reprex) 
#library(datapasta)
```

The first package (Reprex) by Jenny Bryan means "*Repr*oduceable *Ex*ample" and what it does is essentially the code you copy will be pasted - from the clipboard - in a markdown chunk along with the output results (including the errors - if any). This is so that it can be easily copied and pasted for the person helping you. The caveat is that the code has to be "self contained" and be able to run everything it needs to reproduce the error (or result). So this means you have to include all the library(...) code, loading the data (see next paragraph), along with the associated code you want to inspect.

the second package (datapasta) by Miles McBain is the ability to very simply render your data so that it is *significantly easier* to copy and paste for your reprex.  I had *TONS* of trouble doing this originally, it was very diffcult to render my data such that it is easily reproducible. But now thanks to datapasta it is easily done with several lines of code, see below for an example:

```{r datapasta example}
#TD <- head(transitdataclean)
#tribble_paste(TD)
```
There are even more useful commands but i found this absolutely wonderful when trying to copy and paste my data for a reprex or even other applications!

here is Miles McBain's github for datapasta:
<https://github.com/MilesMcBain/datapasta#getting-data-into-source>

### lubridate and hms seem to be incompatible depending what you load first

taken from github issues -- credit to vspinu

*Naming conflicts are common in R's world. In case lubridate is loaded first, you either have to make an alias like lhms <- lubridate::hms or use explicit lubridate::hms in your code. Nothing to be done about this at the moment. hms in lubridate is a parser producing periods, in hms it's a constructor producing durations.*
