---
title: "WpgTransitPassUp_Extracode"
author: "Marcello Nesca"
date: "31/10/2019"
output: html_document
---

```{r loading packages, eval=FALSE}
library(here) 
library(tidyverse)
library(stringr)
library(hms)
library(knitr)
library(reprex)
library(datapasta)
opts_chunk$set(fig.width = 12, fig.height = 8)
```

Ref: <https://data.winnipeg.ca/Transit/Transit-Pass-ups/mer2-irmb>

```{r loading data, eval=FALSE}
transitdata <- read_csv(here::here("data", "Transit_Pass-ups.csv"))
summary(transitdata)
```

### Changed longitude and latitude
For potential spatial analysis in the future
```{r Separating variables!, eval=FALSE}
transitdataclean <- transitdataclean %>%
  separate(Location, into = c("Extra", "Lat1", "Long1"), sep = " ") %>%
  separate(Lat1, into = c("Extra1", "Latitude"), sep = 1) %>%
  separate(Long1, into = c("Longitude", "Extra2"), sep = -1) %>%
  select(-Extra, -Extra1, -Extra2)
```

## checking normality
+ checking normality assumption with inclusion filters
```{r fig.width = 8.5, fig.height = 5}
Q1 %>%
  ggplot(aes(hms(timetopassup))) +
    geom_histogram() +
    xlab("distribution of passups") +
    ylab("Boxplot of hours to passup")
```

## Inferential Analysis - Question 1..cont..
* Is time to pass-up statistically significant between major bus routes to UofM between 2016-2018 for Full Bus Pass-ups? (less time is worse!)
  + checking equal variances
```{r}
leveneTest(timetopassup ~ RouteNumber, data = Q1)
```

## Inferential Analysis - Question 1..cont..
* Is time to pass-up statistically significant between major bus routes to UofM between 2016-2018 for Full Bus Pass-ups? (less time is worse!)
```{r}
resultQ1 <- aov(timetopassup ~ RouteNumber, data = Q1)
summary(resultQ1)
TukeyHSD(resultQ1)
```
